The Gravettian-Solutrean transition in westernmost Iberia: new data from the sites of Vale Boi and Lapa do Picareiro

### 2020-04-23 analysis/paper/paper.Rmd
- Supplementary material rmd - Added images and tables to supplementary
Paper and supplementary changed - Bibliography solved on paper.rmd - Minor changes to text and spelling - References to figures, tables and supplementary solved
- Minor changes to rmd
- Added table with dates
- Spacing and floats solved

The trantion from the Gravettian to the Solutrean, impacted by adverse climatic conditions during the Heinrich Event 2 (HE2) and the onset of the Last Glacial Maximum (LGM), continues to be an essential topic to understand the Upper Paleolithic evolution and human adaptations to climatic changes at the end of the Pleistocene in Westernmost Europe.

@@

This variability has been interpreted as a technological race in response to the environmental modifications taking course during the HE 2 [@cascalheiraandbicho2013]. The authors suggest that, in order to correspond to the external pressures, there may have been the need to diversify the economic strategies in use until that moment. The high exploitation of quartz, for example, formerly a secondary raw material, and the use of similar reduction strategies between quartz and chert, might represent one such economic response. The same possibly applies to the use of unprecedented raw materials, as is the case of the Proto-Solutrean levels in Vale Boi, where there is the presence of previously unknown raw materials, used mainly for the manufacture of Vale Comprido points [@belmiro2017; @marreiros2009]. Likewise, alterations recorded in territoriality patterns for the Proto-Solutrean can be interpreted has extensive regional networks and may be related to modifications in the landscape that occurred during the HE 2 [@cascalheiraandbicho2013; @cascalheira2019].

@@

By understanding the need to better comprehend the Proto-Solutrean and how abrupt climate change impacted these communities, this study's presents the results of the analysis of stone tool assemblages coming from levels attributed to the Gravettian-Solutrean transition timeframe, from two recently excavated sites - Vale Boi (southern Portugal) and Lapa do Picareiro (central Portugal).

Specificallt, by combining techno-typological information with stratigraphy and absolute chronology results, we present new data on: 1) the technological organization of levels 5/4E of Vale Boi, and levels U/T from Lapa do Picareiro; 2) a stratigraphical and chronological separation between Terminal Gravettian horizons and Proto-Solutrean occupations at both sites, revealing the application of the same general set of pattern to geographic areas outside of the Estremadura; 3) and, consequently, the validity of the three-stage model as the best suited for the currently available data.

@@

For lithic analysis we adopted an attribute-based methodology, which aims to describe morphological and metrical attributes of technological classes. The attributes analyzed followed those present in specialized literature, such as @brezillon1968 and @tixier1980, paired with apprcascalheira2010,hes used in o)ther Upper Paleolithic lithic attribute analysis (see e.g. @bicho1992, @zilhao1997, @cascalheira2010, @almeida2000).

@@

Data collection was done through E4, a software developed by Dibble and McPherron (2003). The complete Data Dictionary with all variables, attributes, values and description, with reference to the consulted literature is available in our research compendium at https://doi.org/xxxx.

After data collection, the databases were imported into R environment, where the information was processed through the creation of descriptive statistical analysis and writing of this article. We include the entire R code used for all the analysis and visualizations contained in this paper is in our online research compendium. To produce those files, we followed the procedures described by Marwick et al. (XXXX) for the creation of research compendiums to enhance the reproducibility of research. To enable maximum re-use, our code is released under the MIT license, our data as CC-0, and our figures as CC-BY (for more information, see Marwick XXXX).

@@

Vale Boi is an open-air site and rockshelter, located on the western coast of Algarve (Portugal), near a small homonymous village, within the municipality of Vila do Bispo (Figure \@ref(fig:protomap)). The site is situated in a small valley that runs south to the Atlantic coast, about 2 km distance, relatively open, with a natural boundary to the east and bordered by a limestone hill through all its extension. This hill is marked, at specific points, by limestone exposures that form rock shelters with faces facing west or southwest (Figure \@ref(fig:vbphoto))[@bicho2003; @cascalheira2010; @cascalheiraetal2008].

@@

Since 2012, six layers have been identified (Table \@ref(tab:vbstrat)). In some of these layers, lateral sediment variations were identified, which were coded in the field through the concatenation of a letter to the layer number (*e.g.* 4E). In many of these cases, the isolation of this vertical variation did not show any patterns in terms of spatial concentration of materials, although in others, like the 4E facies, the subdivision of the layer correlated with the spatial distribution of Vale Comprido technology.

@@

These layers, 4E and 5, are vertically contiguous, separated by a relatively flat surface covered by large limestone blocks. Unlike what was expected, however, these levels did not reveal a Gravettian assemblage with characteristics similar to previous years [@marreiros2009], but rather materials with patterns similar to those expected in a Proto-Solutrean assemblage (the most noticeable of all being the Vale Comprido points and blanks). The context was dated using a charcoal sample to c. 24.7-25.3 ka cal BP, and a shell sample to c. 23.7-24.1 ka cal BP (Table \@ref(tab:protodates)).

@@

@@

One of the most noticeable features about the vertical distribution of pieces in the Terrace (row H and I; Figure \@ref(fig:spatialvbfig)) is the vertical constraint in the distribution of dolerite pieces, mostly concentrated on Layer 4E and upper levels of Layer 5, coinciding also with the higher concentration of lithic materials in these layers. Likewise, the presence of Vale Comprido points is also associated with Layer 4E and the top levels of Layer 5 [@belmiro2017], which seems to occur around 25 ka cal BP (Figure \@ref(fig:c14VB)). The WK-44416 date seems to show that, at least from around c. 26 ka cal BP, the site was occupied even if with lower intensity, a date that falls onto the Final Gravettian/Proto-Solutrean or Final Gravettian/Terminal Gravettian transition in the traditional models [@zilhao1997].

@@

A total of 26,703 pieces were analysed for both groups, 11,094 from Lower 5 group (Table \@ref(tab:general1)) and 15,609 for the Upper 5/4E (Table \@ref(tab:general2)). Most of these are chips (70.62% from Lower 5 and 65.68% from Upper 5/4E), followed by shatters, which make up c. 21% and c. 24% of the groups, respectively. These extremely high numbers for debris are mostly the result of quartz use, which can be explained not only by on-site knapping of this raw material but mostly by its breakage patterns (especially when coarser), which typically produces more waste than in other raw materials.

@@

Most of the analysed core platforms are plain or cortical. On Upper 5/4E there is a small frequency of faceted platforms on both quartz and chert (3.4% and 18.5% respectively). Platforms width and thickness means show smaller platforms for chert on both phases, while greywacke shows the highest means for platform measurements (Tables \@ref(tab:coremetricsVB1) and \@ref(tab:coremetricsVB2), Appendix B). This pattern is similar for other measurements, for which chert and quartz exhibit the smaller means.

Regarding flakes, morphologically, Lower 5 flakes seem to have mostly irregular shapes, with frequencies over 45% for most raw materials, followed by convergent and parallel shapes in much lower frequencies (under 15%) (Table \@ref(tab:flakeattributeVB1), Appendix B). Dolerite, unlike other raw materials, shows only convergent and parallel shapes even if with low representativity (n=3). Cross sections are mostly irregular and triangular, varying slightly in frequency regarding each raw material. All raw materials, with exception of quartz, show high frequencies of straight profiles (over 40%), chert showing a similar percentage between straight and curved profiles (44.3% and 42.9% respectively). Quartz, however, has a frequency of 65% of straight profiles, showing much smaller frequencies for other types.

On Upper 5/4E (Table \@ref(tab:flakeattributeVB2), Appendix B), irregular shapes are still frequent, representing over 25% on all raw materials. Convergent and parallel shapes show higher frequencies (over 20%) for quartz, chert and dolerite. Alike Lower 5, cross sections are mostly irregular (over 35% for most raw materials) or triangular (over 25% for all raw materials), and profile type show the same patterns and similar percentages as Lower 5.

@@

As seen before, retouched pieces make up a small percentage of the assemblage, with 167 identified pieces, from which 55 can be inserted in the Lower 5 group (Table \@ref(tab:general1)), while 112 in the Upper 5/4E group (Table \@ref(tab:general2)).

Lower 5 is comprised of several types, from which a few stand out by their high numbers, such as retouched flakes, corresponding to c. 27% of all retouched pieces, splintered pieces, with c. 25% of representativity, followed by notches with a frequency of 12.7%, the latter two occurring in similar numbers in chert (n=6 and n=4, respectively) and quartz (n=8 and n=3) (Table \@ref(tab:retouchVB1)).

@@

For Upper 5/4E, splintered pieces are still very frequent (c. 24%), in both chert and quartz, followed by endscrapers, which represent c. 22% of the total retouched pieces assemblage. Most of the endscapers are chert, although it may be relevant to point out that two of them are made of dolerite. Notches are, once again, the third most frequent retouched typology, representing c. 12%. Every other retouched tool type has frequencies under 10% (Table \@ref(tab:retouchVB2)).

@@

The interior of the chamber is about 11 x 14 m, and has more than 10 m of coarse sedimentary infill in inclined beds, derived from roof collapse, gravity flows and fine sediment infiltration (Figure \@ref(fig:lpcave)) [@benedettietal2019]. The cave opening is marked by the existence of a large limestone cone of stone blocks [@bicho2006] while the sediment inside the cave consists of smaller and angular limestone clasts in a matrix of fine sediment [@benedettietal2019].

```{r lpcave, fig.cap="Lapa do Picareiro. Generalized cross section showing surface topography, shape of the cave, and area excavated into sedimentary fill. After Benedetti et al. (2019)."}

knitr::include_graphics(here::here("analysis/figures/lp_cross_section.jpg"))

@@

Table \@ref(tab:lpstrat) (Supplementary material), adapted from Benedetti et. al [-@benedettietal2019], shows the complete description of all currently identified levels in the site and, whenever existent, associated lithic assemblages. From the 34 levels described in the table, 23 show human occupations or association with a lithic assemblage, and 20 of these can be attributed to the Upper Paleolithic:

@@

Layer T/U is about 15 cm thick, with the presence of possible combustion features and distinct activity areas. The recovered artifacts show a rich faunal assemblage, with a high presence of red deer, though ibex presence is also important, and a superabundance of rabbit bones. A few bone tools were also found. Other recovered artifacts also include a perforated marine shell and a perforated red deer canine tooth (Figure \@ref(fig:teeth)).

The niche area was where most of the lithic assemblage from these levels was found. There is an apparent concentration of lithics, comprised mostly of quartz and rock crystal, and some chert, all raw materials appearing mostly in the shape of bladelets (Figure \@ref(fig:spatialut)).

For U and lower T, radiocarbon dates (Figure \@ref(fig:c14LP)), all obtained from charcoal samples, show results of c. 27-26 cal ka BP, except for one of lower T dates which presents a result of c. 25-24 cal ka BP.

@@

There is a clear separation between the two groups, one composed of levels U and lower T, and the other of middle T. This is also visible on the spatial dispersion of lithics and other artifacts (Figure \@ref(fig:spatialut) supplementary material), with an accumulation on the bottom left, correspondent to the U/T levels, and another aggregation in the middle portion, to the right, ranging from 20-40 cm of depth difference between the groups.

@@

A total of 376 pieces were analysed, 196 coming from the U/Lower T phase (Table \@ref(tab:general1lp)) and 180 from the Middle T group (Table \@ref(tab:general2lp)). In both groups, debitage waste is mostly composed of chips, which represents 49.5% of the U/Lower T group and 37.2% of the Middle T group. As in Vale Boi, these values for chippage can be mostly explained by quartz breakage patterns.

@@

Single platform, prismatic and pyramidal cores were identified in the Middle T group (Table \@ref(tab:coreattributesLP2), Appendix B). These were mainly used to remove flakes and only prismatic cores showed evidence for the production of mixed blanks. Platforms tend to be unfacetted, although a small number of cores presented dihedral platforms (2 in Middle T, in quartz and chert). The number of debitage surfaces is greater on chert cores for the Middle T group, while quartz varies between single and three faces.

@@

On the U/Lower T phase, for quartz, there seems to be the predominance of parallel shapes (35%) (Table \@ref(tab:flakeattributesLP1), Appendix B), followed by convergent ones (30%). Chert, however, shows higher percentages for irregular (28.6%) and divergent (42.9%) shapes. Cross sections are mostly triangular for all raw materials (over 40%), although chert also shows high frequencies of trapezoidal cross sections (42.9%). Profiles are, for all raw materials, mostly straight (60% on quartz and 85.7% on chert) and terminations show relevant raw material differences, with quartz having bigger frequencies of feathered and hinged terminations (35% and 40% respectively) and chert having a frequency of 42.9% on pointed terminations. For all raw materials, there is the predominance of unidirectional dorsal patterns, with high frequencies over 85%. Dorsal scars show higher frequencies for 2 and 3 scars on quartz (35% and 40% respectively), while other raw materials show higher frequencies for 3 dorsal scars only (42.9% on chert).

On the Middle T phase (Table \@ref(tab:flakeattributesLP2), Appendix B), quartz shows higher frequencies for irregular shapes (53.8%), followed by convergent ones (38.5%). Contrary to the patterns seen on the previous group, chert shows higher frequencies for parallel shapes (40.6%). For all raw materials, cross sections are predominantly triangular (over 40%) with straight profiles (over 50%) and feathered terminations for chert (40.6%), whereas quartz shows higher values for hinged terminations (46.2%) followed by feathered (38.5%). Alike the U/Lower T group, there is the predominance of unidirectional dorsal scar patterns (over 90% for all raw materials), with high frequencies for 2 dorsal scars in quartz (61.5%) and 2-3 dorsal scars in chert (34.4% both).

@@

In fact, both blades and bladelets show a majority of parallel and convergent edge shapes, straight and curved profiles. Dorsal scars are unidirectional on the dorsal surfaces of all artifacts, with a majority of two previous removals for quartz and two to four scars for chert. Platforms are mostly plain, with no cortex (Tables \@ref(tab:elongtableLP1) and \@ref(tab:elongtableLP2), Appendix B).

@@

For the Middle T group (Table \@ref(tab:retouchLP2)), there is a smaller variety of retouched pieces, all in chert: a notch, two retouched flakes, one truncation, and possibly a Vale Comprido point, although the thinning retouch at the platform questionable.

@@

### Suplemmentary

```{r general1, echo=FALSE}

### General table Lower 5
db_phase1 <- dataset %>% 
  filter(Phase=="Lower 5")

table1 <- db_phase1 %>% 
  select(RawMaterial, Class) %>% 
  filter(Class != "Chip") %>% 
  dplyr::group_by(RawMaterial)

chip_table1 <- db_phase1 %>% 
  filter(Class == "Chip") %>%
  select(RawMaterial, ChipQuantity) %>% 
  na.omit() %>% 
  dplyr::group_by(RawMaterial) %>% 
  dplyr::summarise(ChipQuantity = sum(ChipQuantity)) %>% 
  tidyr::spread(RawMaterial, ChipQuantity) %>% 
  mutate(Class = "Chip") %>% 
  select(Class, Quartz, Chert, Greywacke, 
         Dolerite, Chalcedony, Other)

general_table1 <- table(table1$Class, table1$RawMaterial)
general_table1 <- as.data.frame.matrix(general_table1)
general_table1 <- general_table1 %>% 
  select(Quartz, Chert, Greywacke, 
         Dolerite, Chalcedony, Other) %>% 
  tibble::rownames_to_column()
general_table1 <- dplyr::rename(general_table1, "Class" = "rowname")
general_table1 <- union(general_table1, chip_table1)

general_table1 <- general_table1 %>% 
  mutate("Quartz (%)" = paste0(round(100 * Quartz/sum(Quartz), 2), "%")) %>% 
  mutate("Chert (%)" = paste0(round(100 * Chert/sum(Chert), 2), "%")) %>% 
  mutate("Greywacke (%)" = paste0(round(100 * Greywacke/sum(Greywacke), 2), "%")) %>% 
  mutate("Dolerite (%)" = paste0(round(100 * Dolerite/sum(Dolerite), 2), "%")) %>% 
  mutate("Chalcedony (%)" = paste0(round(100 * Chalcedony/sum(Chalcedony), 2), "%")) %>% 
  mutate("Other (%)" = paste0(round(100 * Other/sum(Other), 2), "%")) %>% 
  select("Class", "Quartz", "Quartz (%)", "Chert", "Chert (%)", "Greywacke", "Greywacke (%)", "Dolerite",
         "Dolerite (%)", "Chalcedony", "Chalcedony (%)", "Other", "Other (%)") %>% 
  dplyr::rename("Quartz (n)" = "Quartz", "Chert (n)" = "Chert", "Greywacke (n)" = "Greywacke", "Dolerite (n)" = "Dolerite",
                "Chalcedony (n)" = "Chalcedony", "Other (n)" = "Other") %>% 
  adorn_totals(where = "col", name = "Total") %>% 
  mutate("Total (%)" = paste0(round(100 * Total/sum(Total), 2), "%")) %>% 
  adorn_totals(where = "row", name = "Total")

knitr::kable(head(general_table1[1:15], 12), booktabs=TRUE, caption="Vale Boi - Lower 5. Technological class by raw material.")

```{r general2, fig.cap="General table Upper 5/4E.", echo=FALSE}

### General table for Upper 5/4E

db_phase2 <- dataset %>% 
  filter(Phase=="Upper 5/4E")

table2 <- db_phase2 %>% 
  select(RawMaterial, Class) %>% 
  filter(Class != "Chip") %>% 
  dplyr::group_by(RawMaterial)

chip_table2 <- db_phase2 %>% 
  filter(Class == "Chip") %>%
  select(RawMaterial, ChipQuantity) %>% 
  na.omit() %>% 
  dplyr::group_by(RawMaterial) %>% 
  dplyr::summarise(ChipQuantity = sum(ChipQuantity)) %>% 
  tidyr::spread(RawMaterial, ChipQuantity) %>% 
  mutate(Class = "Chip")
chip_table2$Dolerite=0
chip_table2$Other=0
chip_table2 <- chip_table2 %>%
  select(Class,Quartz, Chert, Greywacke, 
         Dolerite, Chalcedony, Other)

general_table2 <- table(table2$Class, table2$RawMaterial)
general_table2 <- as.data.frame.matrix(general_table2)
general_table2 <- general_table2 %>% 
  select(Quartz, Chert, Greywacke, 
         Dolerite, Chalcedony, Other) %>% 
  tibble::rownames_to_column()
general_table2 <- dplyr::rename(general_table2, "Class" = "rowname")
general_table2 <- union(general_table2, chip_table2)

general_table2 <- general_table2 %>% 
  mutate("Quartz (%)" = paste0(round(100 * Quartz/sum(Quartz), 2), "%")) %>% 
  mutate("Chert (%)" = paste0(round(100 * Chert/sum(Chert), 2), "%")) %>% 
  mutate("Greywacke (%)" = paste0(round(100 * Greywacke/sum(Greywacke), 2), "%")) %>% 
  mutate("Dolerite (%)" = paste0(round(100 * Dolerite/sum(Dolerite), 2), "%")) %>% 
  mutate("Chalcedony (%)" = paste0(round(100 * Chalcedony/sum(Chalcedony), 2), "%")) %>% 
  mutate("Other (%)" = paste0(round(100 * Other/sum(Other), 2), "%")) %>% 
  select("Class", "Quartz", "Quartz (%)", "Chert", "Chert (%)", "Greywacke", "Greywacke (%)", "Dolerite",
         "Dolerite (%)", "Chalcedony", "Chalcedony (%)", "Other", "Other (%)") %>% 
  dplyr::rename("Quartz (n)" = "Quartz", "Chert (n)" = "Chert", "Greywacke (n)" = "Greywacke", "Dolerite (n)" = "Dolerite",
                "Chalcedony (n)" = "Chalcedony", "Other (n)" = "Other") %>% 
  adorn_totals(where = "col", name = "Total") %>% 
  mutate("Total (%)" = paste0(round(100 * Total/sum(Total), 2), "%")) %>% 
  adorn_totals(where = "row", name = "Total")

knitr::kable(head(general_table2[1:15], 12), booktabs=TRUE,caption="Vale Boi - Upper 5/4E. Technological class by raw material.")
```

```{r general1lp, echo=FALSE}

## General table for U/Lower T
db_phaseTG <- datasetlp %>% 
  filter(Phase == "U/Lower T")

table_TG <- db_phaseTG %>% 
  select(RawMaterial, Class) %>% 
  filter(Class != "Chip") %>% 
  dplyr::group_by(RawMaterial)

chip_table_TG <- db_phaseTG %>% 
  filter(Class == "Chip") %>%
  select(RawMaterial, ChipQuantity) %>% 
  na.omit() %>% 
  dplyr::group_by(RawMaterial) %>% 
  dplyr::summarise(ChipQuantity = sum(ChipQuantity)) %>% 
  tidyr::spread(RawMaterial, ChipQuantity) %>% 
  mutate(Class = "Chip") %>% 
  select(Class, Quartz, Chert, Other)

general_table_TG <- table(table_TG$Class, table_TG$RawMaterial)
general_table_TG <- as.data.frame.matrix(general_table_TG)
general_table_TG <- general_table_TG %>% 
  select(Quartz, Chert, Other) %>% 
  tibble::rownames_to_column()
general_table_TG <- dplyr::rename(general_table_TG, "Class" = "rowname")
general_table_TG <- union(general_table_TG, chip_table_TG)

general_table_TG <- general_table_TG %>% 
  mutate("Quartz (%)" = paste0(round(100 * Quartz/sum(Quartz), 2), "%")) %>% 
  mutate("Chert (%)" = paste0(round(100 * Chert/sum(Chert), 2), "%")) %>% 
  mutate("Other (%)" = paste0(round(100 * Other/sum(Other), 2), "%")) %>% 
  select("Class", "Quartz", "Quartz (%)", "Chert", "Chert (%)", "Other", "Other (%)") %>% 
  dplyr::rename("Quartz (n)" = "Quartz", "Chert (n)" = "Chert", "Other (n)" = "Other") %>% 
  adorn_totals(where = "col", name = "Total") %>% 
  mutate("Total (%)" = paste0(round(100 * Total/sum(Total), 2), "%")) %>% 
  adorn_totals(where = "row", name = "Total (RM)")

knitr::kable(head(general_table_TG[1:9], 9), booktabs=TRUE,caption="Lapa do Picareiro - U/Lower T. Technological class by raw material.")

```{r general2lp, echo=FALSE}

### Middle T table
db_phasePR <- datasetlp %>% 
  filter(Phase == "Middle T")

table_PR <- db_phasePR %>% 
  select(RawMaterial, Class) %>% 
  filter(Class != "Chip") %>% 
  dplyr::group_by(RawMaterial)

chip_table_PR <- db_phasePR %>% 
  filter(Class == "Chip") %>%
  select(RawMaterial, ChipQuantity) %>% 
  na.omit() %>% 
  dplyr::group_by(RawMaterial) %>% 
  dplyr::summarise(ChipQuantity = sum(ChipQuantity)) %>% 
  tidyr::spread(RawMaterial, ChipQuantity) %>% 
  mutate(Class = "Chip") %>% 
  select(Class,Quartz, Chert, Other)

general_table_PR <- table(table_PR$Class, table_PR$RawMaterial)
general_table_PR <- as.data.frame.matrix(general_table_PR)
general_table_PR <- general_table_PR %>% 
  select(Quartz, Chert, Other) %>% 
  tibble::rownames_to_column()
general_table_PR <- dplyr::rename(general_table_PR, "Class" = "rowname")
general_table_PR <- union(general_table_PR, chip_table_PR)

general_table_PR <- general_table_PR %>% 
  mutate("Quartz (%)" = paste0(round(100 * Quartz/sum(Quartz), 2), "%")) %>% 
  mutate("Chert (%)" = paste0(round(100 * Chert/sum(Chert), 2), "%")) %>% 
  mutate("Other (%)" = paste0(round(100 * Other/sum(Other), 2), "%")) %>% 
  select("Class", "Quartz", "Quartz (%)", "Chert", "Chert (%)", "Other", "Other (%)") %>% 
  dplyr::rename("Quartz (n)" = "Quartz", "Chert (n)" = "Chert", "Other (n)" = "Other") %>% 
  adorn_totals(where = "col", name = "Total") %>% 
  mutate("Total (%)" = paste0(round(100 * Total/sum(Total), 2), "%")) %>% 
  adorn_totals(where = "row", name = "Total (RM)")

>>>>

The transition from the Gravettian to the Solutrean, impacted by adverse climatic conditions during the Heinrich Event 2 (HE2) and the onset of the Last Glacial Maximum (LGM), continues to be an essential topic to understand the Upper Paleolithic evolution and human adaptations to climatic changes at the end of the Pleistocene in Westernmost Europe.

@@

This variability has been interpreted as a technological race in response to the environmental modifications taking course during the HE 2 [@cascalheiraandbicho2013]. The authors suggest that, in order to correspond to the external pressures, there may have been the need to diversify the economic strategies in use until that moment. The high exploitation of quartz, for example, formerly a secondary raw material, and the use of similar reduction strategies between quartz and chert, might represent one such economic response. The same possibly applies to the use of unprecedented raw materials, as is the case of the Proto-Solutrean levels in Vale Boi, where there is the presence of a previously unknown raw material, used mainly for the manufacture of Vale Comprido points [@belmiro2017; @marreiros2009]. Likewise, alterations recorded in territoriality patterns for the Proto-Solutrean can be interpreted has extensive regional networks and may be related to modifications in the landscape that occurred during the HE 2 [@cascalheiraandbicho2013; @cascalheira2019].

@@

By understanding the need to better comprehend the Proto-Solutrean and how abrupt climate change impacted these communities, this study presents the results of the analysis of stone tool assemblages coming from levels attributed to the Gravettian-Solutrean transition timeframe, from two recently excavated sites - Vale Boi (southern Portugal) and Lapa do Picareiro (central Portugal).

Specifically, by combining techno-typological information with stratigraphy and absolute chronology results, we present new data on: 1) the technological organization of levels 5/4E of Vale Boi, and levels U/T from Lapa do Picareiro; 2) a stratigraphical and chronological separation between Terminal Gravettian horizons and Proto-Solutrean occupations at both sites, revealing the application of the same general set of pattern to geographic areas outside of the Estremadura; 3) and, consequently, the validity of the three-stage model as the best suited for the currently available data.

@@

For the lithic analysis we adopted an attribute-based methodology, which aims to describe morphological and metrical attributes of technological classes. The attributes analyzed followed those present in specialized literature, such as @brezillon1968 and @tixier1980, paired with other Upper Paleolithic lithic attribute analyses (see e.g. @bicho1992, @zilhao1997, @cascalheira2010, @almeida2000).

@@

Data collection was done through E4, a software developed by Dibble and McPherron (2003). The complete Data Dictionary with all variables, attributes, values and description, with reference to the consulted literature is available in our research compendium at https://doi.org/xxxx, and can be consulted in the supplementary material (Supplementary table 1).

After data collection, the databases were imported into R environment, where the information was processed through the creation of descriptive statistical analysis and writing of this article. We include the entire R code used for all the analysis and visualizations contained in this paper is in our online research compendium. To produce those files, we followed the procedures described by Marwick et al. [-@marwick2017] for the creation of research compendiums to enhance the reproducibility of research. To enable maximum re-use, our code is released under the MIT license, our data as CC-0, and our figures as CC-BY (for more information, see @marwick2018]. All result tables may also be found in the Supplementary Materials.

@@

Vale Boi is an open-air site and rockshelter, located on the western coast of Algarve (Portugal), near a small homonymous village, within the municipality of Vila do Bispo (in the Supplementary Material). The site is situated in a small valley that runs south to the Atlantic coast, about 2 km distance, relatively open, with a natural boundary to the east and bordered by a limestone hill through all its extension. This hill is marked, at specific points, by limestone exposures that form rock shelters with faces facing west or southwest [@bicho2003; @cascalheira2010; @cascalheiraetal2008].

@@

Since 2012, six layers have been identified (Supplementary table 2). In some of these layers, lateral sediment variations were identified, which were coded in the field through the concatenation of a letter to the layer number (*e.g.* 4E). In many of these cases, the isolation of this vertical variation did not show any patterns in terms of spatial concentration of materials, although in others, like the 4E facies, the subdivision of the layer correlated with the spatial distribution of Vale Comprido technology.

@@

These layers, 4E and 5, are vertically contiguous, separated by a relatively flat surface covered by large limestone blocks. Unlike what was expected, however, these levels did not reveal a Gravettian assemblage with characteristics similar to previous years [@marreiros2009], but rather materials with patterns similar to those expected in a Proto-Solutrean assemblage (the most noticeable of all being the Vale Comprido points and blanks). The context was dated using a charcoal sample to c. 24.7-25.3 ka cal BP, and a shell sample to c. 23.7-24.1 ka cal BP (complete table with calibrated radiocarbon dates may be found in the Supplementary Materials).

@@

One of the most noticeable features about the vertical distribution of pieces in the Terrace (row H and I; Supplementary Figure 4) is the vertical constraint in the distribution of dolerite pieces, mostly concentrated on Layer 4E and upper levels of Layer 5, coinciding also with the higher concentration of lithic materials in these layers. Likewise, the presence of Vale Comprido points is also associated with Layer 4E and the top levels of Layer 5 [@belmiro2017], which seems to occur around 25 ka cal BP (Figure \@ref(fig:c14VB)). The WK-44416 date seems to show that, at least from around c. 26 ka cal BP, the site was occupied even if with lower intensity, a date that falls onto the Final Gravettian/Proto-Solutrean or Final Gravettian/Terminal Gravettian transition in the traditional models [@zilhao1997].

@@

@@

A total of 26,703 pieces were analysed for both groups, 11,094 from Lower 5 group and 15,609 for the Upper 5/4E (both tables may be found in the Supplementary Materials). Most of these are chips (70.62% from Lower 5 and 65.68% from Upper 5/4E), followed by shatters, which make up c. 21% and c. 24% of the groups, respectively. These extremely high numbers for debris are mostly the result of quartz use, which can be explained not only by on-site knapping of this raw material but mostly by its breakage patterns (especially when coarser), which typically produces more waste than in other raw materials.

@@

Most of the analysed core platforms are plain or cortical. On Upper 5/4E there is a small frequency of faceted platforms on both quartz and chert (3.4% and 18.5% respectively). Platforms width and thickness means show smaller platforms for chert on both phases, while greywacke shows the highest means for platform measurements. This pattern is similar for other measurements, for which chert and quartz exhibit the smaller means.

Regarding flakes, morphologically, Lower 5 flakes seem to have mostly irregular shapes, with frequencies over 45% for most raw materials, followed by convergent and parallel shapes in much lower frequencies (under 15%). Dolerite, unlike other raw materials, shows only convergent and parallel shapes even if with low representativity (n=3). Cross sections are mostly irregular and triangular, varying slightly in frequency regarding each raw material. All raw materials, with exception of quartz, show high frequencies of straight profiles (over 40%), chert showing a similar percentage between straight and curved profiles (44.3% and 42.9% respectively). Quartz, however, has a frequency of 65% of straight profiles, showing much smaller frequencies for other types.

On Upper 5/4E, irregular shapes are still frequent, representing over 25% on all raw materials. Convergent and parallel shapes show higher frequencies (over 20%) for quartz, chert and dolerite. Alike Lower 5, cross sections are mostly irregular (over 35% for most raw materials) or triangular (over 25% for all raw materials), and profile type show the same patterns and similar percentages as Lower 5.

@@

As seen before, retouched pieces make up a small percentage of the assemblage, with 167 identified pieces, from which 55 can be inserted in the Lower 5 group, while 112 in the Upper 5/4E group.

Lower 5 is comprised of several types, from which a few stand out by their high numbers, such as retouched flakes, corresponding to c. 27% of all retouched pieces, splintered pieces, with c. 25% of representativity, followed by notches with a frequency of 12.7%, the latter two occurring in similar numbers in chert (n=6 and n=4, respectively) and quartz (n=8 and n=3).

@@

For Upper 5/4E, splintered pieces are still very frequent (c. 24%), in both chert and quartz, followed by endscrapers, which represent c. 22% of the total retouched pieces assemblage. Most of the endscapers are chert, although it may be relevant to point out that two of them are made of dolerite. Notches are, once again, the third most frequent retouched typology, representing c. 12%. Every other retouched tool type has frequencies under 10%.

@@

The interior of the chamber is about 11 x 14 m, and has more than 10 m of coarse sedimentary infill in inclined beds, derived from roof collapse, gravity flows and fine sediment infiltration (Supplementary Figure 9) [@benedettietal2019]. The cave opening is marked by the existence of a large limestone cone of stone blocks [@bicho2006] while the sediment inside the cave consists of smaller and angular limestone clasts in a matrix of fine sediment [@benedettietal2019].

@@

A complete description of all currently identified levels in the site and, whenever existent, associated lithic assemblages, may be found in the Supplementary Materials. From the 34 levels described in the table, 23 show human occupations or association with a lithic assemblage, and 20 of these can be attributed to the Upper Paleolithic:

@@

Layer T/U is about 15 cm thick, with the presence of possible combustion features and distinct activity areas. The recovered artifacts show a rich faunal assemblage, with a high presence of red deer, though ibex presence is also important, and a superabundance of rabbit bones. A few bone tools were also found. Other recovered artifacts also include a perforated marine shell and a perforated red deer canine tooth.

The niche area was where most of the lithic assemblage from these levels was found. There is an apparent concentration of lithics, comprised mostly of quartz and rock crystal, and some chert, all raw materials appearing mostly in the shape of bladelets.

For U and lower T, radiocarbon dates (Supplementary Table 3), all obtained from charcoal samples, show results of c. 27-26 cal ka BP, except for one of lower T dates which presents a result of c. 25-24 cal ka BP.

@@

There is a clear separation between the two groups, one composed of levels U and lower T, and the other of middle T. This is also visible on the spatial dispersion of lithics and other artifacts (Supplementary Figure 11), with an accumulation on the bottom left, correspondent to the U/T levels, and another aggregation in the middle portion, to the right, ranging from 20-40 cm of depth difference between the groups.

@@

A total of 376 pieces were analysed, 196 coming from the U/Lower T phase and 180 from the Middle T group. In both groups, debitage waste is mostly composed of chips, which represents 49.5% of the U/Lower T group and 37.2% of the Middle T group. As in Vale Boi, these values for chippage can be mostly explained by quartz breakage patterns.

@@

Single platform, prismatic and pyramidal cores were identified in the Middle T group. These were mainly used to remove flakes and only prismatic cores showed evidence for the production of mixed blanks. Platforms tend to be unfacetted, although a small number of cores presented dihedral platforms (2 in Middle T, in quartz and chert). The number of debitage surfaces is greater on chert cores for the Middle T group, while quartz varies between single and three faces.

@@

On the U/Lower T phase, for quartz, there seems to be the predominance of parallel shapes (35%), followed by convergent ones (30%). Chert, however, shows higher percentages for irregular (28.6%) and divergent (42.9%) shapes. Cross sections are mostly triangular for all raw materials (over 40%), although chert also shows high frequencies of trapezoidal cross sections (42.9%). Profiles are, for all raw materials, mostly straight (60% on quartz and 85.7% on chert) and terminations show relevant raw material differences, with quartz having bigger frequencies of feathered and hinged terminations (35% and 40% respectively) and chert having a frequency of 42.9% on pointed terminations. For all raw materials, there is the predominance of unidirectional dorsal patterns, with high frequencies over 85%. Dorsal scars show higher frequencies for 2 and 3 scars on quartz (35% and 40% respectively), while other raw materials show higher frequencies for 3 dorsal scars only (42.9% on chert).

On the Middle T phase, quartz shows higher frequencies for irregular shapes (53.8%), followed by convergent ones (38.5%). Contrary to the patterns seen on the previous group, chert shows higher frequencies for parallel shapes (40.6%). For all raw materials, cross sections are predominantly triangular (over 40%) with straight profiles (over 50%) and feathered terminations for chert (40.6%), whereas quartz shows higher values for hinged terminations (46.2%) followed by feathered (38.5%). Alike the U/Lower T group, there is the predominance of unidirectional dorsal scar patterns (over 90% for all raw materials), with high frequencies for 2 dorsal scars in quartz (61.5%) and 2-3 dorsal scars in chert (34.4% both).

@@

In fact, both blades and bladelets show a majority of parallel and convergent edge shapes, straight and curved profiles. Dorsal scars are unidirectional on the dorsal surfaces of all artifacts, with a majority of two previous removals for quartz and two to four scars for chert. Platforms are mostly plain, with no cortex.

@@

For the Middle T group, there is a smaller variety of retouched pieces, all in chert: a notch, two retouched flakes, one truncation, and possibly a Vale Comprido point, although the thinning retouch at the platform questionable.

@@
